---
title: "GDPCaseStudy"
author: "Armand Post"
date: "June 12, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Set working directory (you'll need to update this to your local working directory)
```{r}
setwd("/SMU/Doing Data Science/Unit 6/Case Study")
```

## Load Libraries, install "stringr" package if needed
```{r}
##install.packages("stringr")
library(plyr)
library(downloader)
library(stringr)
```

## Download the datasets
```{r}

download("https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FGDP.csv",destfile="GDPData.csv")
download("https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FEDSTATS_Country.csv",destfile="EducationalData.csv")

cat("Data downloaded into current working directory\n")
```

## Load the data into R
```{r}
GDP <- read.csv("GDPData.csv",skip=4)
Education <- read.csv("EducationalData.csv",na.strings=c("", "NA"))
```

## Assess the data
```{r}
##head(GDP)
##head(Education)
summary(GDP)
##summary(Education)
```

## Assign header names for GDP, and remove blank columns
```{r}
GDP <- GDP[,-3,-7]
colnames(GDP) <- c('CountryCode','GDPRank','CountryName','GDP')
GDP <- GDP[,-5,-6]
GDP <- GDP[,-5,-6]
GDP <- GDP[,-5,-6]
GDP <- GDP[,-5,-6]
GDP <- GDP[,-5,-6]
head(GDP)
```

## From the GDP Summary, it was found that 24 values for GDP are missing.  Those values are removed below.  Also remove blank rows
```{r}
GDP <- subset(GDP, GDP != "..")
GDP <- subset(GDP, CountryName != "Djibouti")
GDP <- subset(GDP, GDP != "")
```


## Merge the Datasets and reduce to only columns needed.
```{r}
FullData <- merge(GDP,Education, by.x = "CountryCode", by.y = "CountryCode", all=TRUE)

GDPRecords <- GDP[,c("CountryCode","CountryName")]
EduRecords <- Education[,c("CountryCode","Short.Name")]
FullData2 <- FullData[,c("CountryCode","GDPRank","CountryName","GDP","Short.Name","Income.Group")]
FullData2 <- FullData2[order(FullData2$GDP),]
```

## Show row counts.  Note that two records were in GDP, but not Education (Andorra, South Sudan).  Additonally, 21 records were in Education, but not GDP
```{r}
nrow(GDPRecords)
nrow(EduRecords)
nrow(FullData2)
```

## Remove the records that didn't match and non country specific data.  Also add in a GDP rank column with leading zeros for sorting
```{r}
FullData3 <- subset(FullData2, GDPRank != "")
FullData3$GDPRank2 <- str_pad(FullData3$GDPRank, width=3, side="left", pad="0")
FullData3 <- FullData3[order(FullData3$GDPRank2),]
nrow(FullData3)
head(FullData3)
```



