---
title: "GDPCaseStudy"
author: "Armand Post"
date: "June 12, 2016"
output:
    html_document:
        keep_md: true
---

## GDP and Education Level Case Study

The purpose of this case study is to assess GDP Rank and GDP in USD using data from the worldbank.org website.  Two datasets are used and merged with the purpose of appending income group with the GDP data.  Since countries with large populations may have a large total GDP, but low GDP per capita, grouping by income group will give a better assessment of a country's relative productivity and citezens' quality of life.

### Deliverable
Since reproducibility is an important aspect of this assignment, the deliverable will be a github repository with R Markdown code, data dictionary, and the original datasets.  The code will be runable with one alteration, changing the working directory to one on the running computer's local drive.  However, there are a number of packages used which may need to be installed.

## The Code
Code chunks were given names and the code was intentionally split into chunks that perform specific functions.  This was with the purpose in mind of making references to the code easier to follow.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Set working directory (you'll need to update this to your local working directory)
```{r working directory}
setwd("/SMU/Doing Data Science/Unit 6/CaseStudy/CaseStudyWeek6")
```

## Load Libraries, install "stringr" package if needed
```{r Libraries}
##install.packages("stringr")
library(plyr)
library(downloader)
library(stringr) # needed to place leading zeros on GDP Rank for order by
library(dplyr)
library(ggplot2)
library(scales) # needed for formatting y-axis labels to non-scientific type
```

## Download the datasets
```{r Download}
download("https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FGDP.csv",destfile="GDPData.csv")
download("https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FEDSTATS_Country.csv",destfile="EducationalData.csv")

cat("Data downloaded into current working directory\n")
```

## Load the data into R
```{r Data Load}
GDP <- read.csv("GDPData.csv",skip=4)
Education <- read.csv("EducationalData.csv",na.strings=c("", "NA"))
```

## Assess the data
```{r Summary Statistics}
##head(GDP)
##head(Education)
summary(GDP)
summary(Education)
```

## Assign header names for GDP, and remove blank columns
```{r Headers}
GDP <- GDP[,-3,-7]
colnames(GDP) <- c('CountryCode','GDPRank','CountryName','GDP')
GDP <- GDP[,-5,-6]
GDP <- GDP[,-5,-6]
GDP <- GDP[,-5,-6]
GDP <- GDP[,-5,-6]
GDP <- GDP[,-5,-6]
head(GDP)
```

## From the GDP Summary, it was found that 24 values for GDP are missing.  Those values are removed below.  Also remove blank rows and NAs
```{r Missing GDP}
GDP <- subset(GDP, GDP != "..")
GDP <- subset(GDP, CountryName != "Djibouti")
GDP <- subset(GDP, GDP != "")
GDP <- na.omit(GDP)
```


## Merge the Datasets and reduce to only columns needed.
```{r Merge}
FullData <- merge(GDP,Education, by.x = "CountryCode", by.y = "CountryCode", all=TRUE)

GDPRecords <- GDP[,c("CountryCode","CountryName")]
EduRecords <- Education[,c("CountryCode","Short.Name")]
FullData2 <- FullData[,c("CountryCode","GDPRank","CountryName","GDP","Short.Name","Income.Group")]
FullData2 <- FullData2[order(FullData2$GDP),]
```

## Question 1.  Show row counts.  Note that two records were in GDP, but not Education (Andorra, South Sudan).  Additonally, 21 records were in Education, but not GDP
```{r Rows}
nrow(GDPRecords)
nrow(EduRecords)
nrow(FullData2)
```

## Remove the records that didn't match and non country specific data.  Also add in a GDP rank column with leading zeros for sorting and sort by descending order.
```{r Order}
FullData3 <- subset(FullData2, GDPRank != "")

FullData3$GDPRank2 <- str_pad(FullData3$GDPRank, width=3, side="left", pad="0")
FullData3 <- FullData3[order(FullData3$GDPRank2, decreasing = TRUE),]
nrow(FullData3)
head(FullData3)
```

## For Question 2, once the data is in descending order, show the 13th from lowest GDP country.
#### From the below, we can see that Grenada and St. Kitts and Nevis are tied for 13th from last.
```{r Show 13}
head(FullData3,13)
```

## For Question 3, What are the average GDP rankings for the "High income: OECD" and "High income: non OECD" groups?
```{r Average GDP}
## Convert GDP to numeric
FullData3$GDP = gsub(pattern = ",", replacement = "", FullData3$GDP)
FullData3[, c(4)] <- sapply(FullData3[, c(4)], as.numeric)
ddply(FullData3, .(Income.Group), summarize, GDPMean = mean(GDP))
head(FullData3)
```
### High income: nonOECD countries have a mean GDP of $104,349 while High income: OECD countries have a mean income of $1,483,917

## Question 4.  Plot the GDP data for all countries
```{r Plot}
scale_y_continuous(labels=function(n){format(n, scientific = FALSE)})
Graph1 <- ggplot(FullData3, aes(x = GDPRank2, y = GDP, color = Income.Group))
Graph1 + geom_point() + scale_y_continuous(labels = comma) + geom_point()
```

## Question 4.  Since the data has a few high GDP countries and many low GDP countries, the above graph is difficult to interpret due to the scale being off.  The following step performs a Log Transformation of GDP and replots.
```{r Log Plot}
scale_y_continuous(labels=function(n){format(n, scientific = FALSE)})
Graph1 <- ggplot(FullData3, aes(x = GDPRank2, y = log(GDP), color = Income.Group))
Graph1 + geom_point() + scale_y_continuous(labels = comma) + geom_point()
```

### Once the data is log transformed, it takes on an S-Shape.  We can also see high income: OECD countries make up most of the high GDP Rank countries, while high income: nonOECD countries tend to lag behind their OECD counterparts.  There are some exceptions, most notably China which has a very high GDP rank despite being classified as a lower middle income nation.  Lower middle income countries typically appear around the middle and low end of GDP Rank.

## Cut the GDP rankings into quintile Groups.
```{r Quintiles}
FullData3$Quintile <- with(FullData3, factor(findInterval(GDP, c(-Inf, quantile(GDP, probs=c(0.20, 0.40, 0.60, 0.80)), Inf)), labels=c("Q5","Q4","Q3","Q2","Q1")))
```

## Make a crosstab of Quintiles vs Income Groups
```{r Crosstab}
# Load function
source("http://pcwww.liv.ac.uk/~william/R/crosstab.r")

FullData3$Income.Group2 <- NA

# shorten income group variable names to fit and create order
FullData3$Income.Group2[FullData3$Income.Group == "High income: OECD"] <- "1 HI: OECD"
FullData3$Income.Group2[FullData3$Income.Group == "High income: nonOECD"] <- "2 HI: nOECD"
FullData3$Income.Group2[FullData3$Income.Group == "Upper middle income"] <- "3 UMI"
FullData3$Income.Group2[FullData3$Income.Group == "Lower middle income"] <- "4 LMI"
FullData3$Income.Group2[FullData3$Income.Group == "Low income"] <- "5 LI"
head(FullData3)

# Frequency count
crosstab(FullData3, row.vars = "Quintile", col.vars = "Income.Group2", type = "f")
# Frequency count
crosstab(FullData3, row.vars = "Quintile", col.vars = "Income.Group2", type = "t", percentages = FALSE)
```

## Question 5.  From the above, there are 5 countries that are lower middle income and in the first quintile.
## Also of note are the distributions for the income group categories.  High income: OECD countries tend to fall in the two upper quintiles, while high income: non OECD and Upper Middle Income countries are spread somewhat evenly amongst quintiles.  Lower middle income and Lower income are skewed towards tht lower quartiles.
